---
title: "ANA 515 Assignment 3, Visualizations Activity"
author: "Mel Sparks"
date: "4/17/2022"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(tibble)
library(scales)
library(quantmod)
```

## Greenhouse Gas Emissions Dataset

This dataset contains information regarding how many tons of greenhouse gas emissions are produced annually in the world and by country from 1990 to 2018. I’m hoping to ascertain whether greenhouse gas production has increased or decreased worldwide, as well as if any top producing countries have made progress towards decreasing their annual output.

```{r ghg data prep, include=FALSE} 
#I need to prep this file by including everything I did in assignment 2
ghgemissions <- read.csv("C:/Users/melin/OneDrive/Desktop/Data Analytics M.S/ANA515/Week 4/ALL GHG_historical_emissions.csv")
#I want to drop columns I don't need; all of these columns have the same entry for all observations
ghgdrop <- subset(ghgemissions, select = -c(Data.source, Sector,Gas,Unit))
#Now I'll limit my data to the top producing countries, any who produced less than 1000 tons of ghg emissions in 2018
ghgtopprod <- filter(ghgdrop, X2018>1000)
#Having the World total is useful, but it will skew my summary statistics at the end of this, so I'll put it in it's own dataset
ghgworld <- filter(ghgtopprod, X2018>20000)
ghgtopprod <- filter(ghgtopprod, X2018<20000)
#Apparently X1990 was accepted as a character class, I need to change that to integer
ghgtopprod$X1990 <- as.integer(ghgtopprod$X1990)

```


## This dataframe has `r nrow(ghgtopprod)` rows and `r ncol(ghgtopprod)` columns. Below, we can see a sample of how the dataframe looks.

```{r ghg data original, echo=FALSE} 
#Here's a sample of what the data looks like using the head function
head(ghgtopprod,n=c(5,6))
```

By looking at the dimensions, we can ascertain that the dataframe continues for another 20+ years. This won't work for the type of visualizations I mentioned that we wanted to look at in the intro, so we'll take a few steps to reformat the dataframe.

```{r ghg data reformat 1, include=FALSE} 
#First we need to swap the columns and rows
final_df<- as.data.frame(t(ghgtopprod))
#Then we need to change the column names to what is listed in the first row
colnames(final_df)<-as.character(final_df[1,])
#And then we delete the first row, because it's the header now
final_df <- subset(final_df, China!="China")
#Now we'll change the row names to the first column and call it 'Year'
final_df <- tibble::rownames_to_column(final_df,"Year")
#And then we'll drop the X off the front off all the years that was produced when we imported the original csv and they were column names
final_df$Year <- substr(final_df$Year, 2,5)
#I noticed while playing around with the data that everything was formatted as character; I wanted to change all of this to numeric values
final_df$Year <- as.numeric(final_df$Year)
final_df$China <- as.numeric(final_df$China)
final_df$`United States` <- as.numeric(final_df$`United States`)
final_df$India <- as.numeric(final_df$India)
final_df$`European Union (27)` <- as.numeric(final_df$`European Union (27)`)
final_df$Russia <- as.numeric(final_df$Russia)
final_df$Indonesia <- as.numeric(final_df$Indonesia)
final_df$Brazil <- as.numeric(final_df$Brazil)
final_df$Japan <- as.numeric(final_df$Japan)
#And after making a few plots, I realized that re-ordering the df by year in ascending order might help (I'm not sure if it actually did anything or not)
attach(final_df)
final_df <- final_df[order(Year),]
detach(final_df)
```

## Now the dataframe has `r nrow(final_df)` rows and `r ncol(final_df)` columns. Below, we can see a sample of how the dataframe looks now.

```{r ghg data reformat 2, echo=FALSE} 
#Now we want to show an example of the changes we've made to the df using the head function again
head(final_df,n=c(5,6))
```

# Visualization 1

## First I want to see the trends for the top GHG emission producing countries for the given time period (1990 - 2018).

```{r ghg visualization 1, echo=FALSE}
#This is the visualization I created for this week's discussion post using the baseline R plot functions
plot(final_df$China,type="o", col="red", ylim=c(1000, 12000), 
     xlab="Year (1990-2018)", ylab="Emissions (tons)", main="Annual GHG Emissions")
lines(final_df$`United States`, type="o", col="blue")
lines(final_df$India, type="o", col="orange")
lines(final_df$`European Union (27)`, type="o", col="green")
lines(final_df$Russia, type="o", col="purple")
legend("topleft", inset=0.01, legend=c("China","United States","India","European Union","Russia"),
       col=c("red","blue","orange","green","purple"), lty=1, cex=0.6,
       title="Legend", text.font=4, box.lty = 0, y.intersp = 0.5)
```

In the graph above, we have the top five GHG producing countries/regions in the world. We can see that China has drastically increased it's GHG emissions over the years. All other top producing countries, by comparison, have either mostly stayed the same -- with the exceptions of Russia, who has allegedly somewhat decreased it's production over the years, and India, who has increased steadily, but not by the margin that China has.

# Visualization 2

I'm also interested in the overall production of GHG emissions worldwide. I want to make a simple graph that shows the change in greenhouse gas emissions worldwide from 1990 to 2018.

```{r ghg worldwide, echo=FALSE}
#I've decided to reformat the worldwide cumulative data as well
ww_df<- as.data.frame(t(ghgworld))
#Then we need to change the column names to what is listed in the first row
colnames(ww_df)<-as.character(ww_df[1,])
#And then we delete the first row, because it's the header now
ww_df <- subset(ww_df, World!="World")
#Now we'll change the row names to the first column and call it 'Year'
ww_df <- tibble::rownames_to_column(ww_df,"Year")
#And then we'll drop the X off the front off all the years that was produced when we imported the original csv and they were column names
ww_df$Year <- substr(ww_df$Year, 2,5)
attach(ww_df)
ww_df <- ww_df[order(Year),]
detach(ww_df)
xValue <- 1990:2018
yValue <- as.numeric(ww_df$World)
data <- data.frame(xValue,yValue)
ggplot(data, aes(x=xValue, y=yValue)) +
  geom_line(color="red",linetype=2) +
  ggtitle("Worldwide GHG Emissions 1990-2018") +
  ylab("Emissions (tons)") +
  xlab("Year (1990-2018)")
```

In the graph above, we can see that greenhouse gas emissions have been increasing worldwide since 1990. This isn't entirely surprising -- in fact I expected GHG emissions to have increased over the years. The thing that is somewhat surprising is that this graph makes me wonder who the key contributors are to the increase. We already know that China's production has drastically increased, but I don't think they account for the increase overall.

# Visualization 3

I want another graph that'll show how much each country/region contributes to the overall increase of GHG emissions over the years. Ideally, I would like to make a stacked bar chart that shows the cumulative increase -- starting with just the top 8 producing countries/regions to keep things simple.

```{r barplot data prep, include=FALSE}
#I'm trying to force the data into a shape that'll let me do what I want with it. For the sake of simplicity, I will do just the top producing countries.
#I'm going to start by making separate dfs for each country
ch_df <- subset(final_df, select=c(Year,China))
us_df <- subset(final_df, select=c(Year,`United States`))
indi_df <- subset(final_df, select=c(Year,India))
eu_df <- subset(final_df, select=c(Year,`European Union (27)`))
ru_df <- subset(final_df, select=c(Year,Russia))
indo_df <- subset(final_df, select=c(Year,Indonesia))
br_df <- subset(final_df, select=c(Year,Brazil))
ja_df <- subset(final_df, select=c(Year,Japan))
#I need to add a country label column to all of the dfs
ch_df$Country <- c("China")
us_df$Country <- c("United States")
indi_df$Country <- c("India")
eu_df$Country <- c("European Union")
ru_df$Country <- c("Russia")
indo_df$Country <- c("Indonesia")
br_df$Country <- c("Brazil")
ja_df$Country <- c("Japan")
#Lastly, I'm pretty sure the columns all need to have the same column names, so I want to rename the existing middle columns to "Emissions" instead of the individual country
names(ch_df)[2]<-'Emissions'
names(us_df)[2]<-'Emissions'
names(indi_df)[2]<-'Emissions'
names(eu_df)[2]<-'Emissions'
names(ru_df)[2]<-'Emissions'
names(indo_df)[2]<-'Emissions'
names(br_df)[2]<-'Emissions'
names(ja_df)[2]<-'Emissions'
total <- rbind(ch_df, us_df, indi_df, eu_df, ru_df, indo_df, br_df, ja_df)
attach(total)
total <- total[order(Year),]
detach(total)
as.numeric(total$Year)
as.numeric(total$Emissions)
```

```{r barplot, echo=FALSE}
ggplot(total, aes(fill=Country, x=Year, y=Emissions)) +
  geom_bar(position="stack",stat="identity") +
  ggtitle("Cumulative GHG Emissions from Top Producing Contries 1990-2018") +
  ylab("Emissions (tons)") +
  xlab("Year (1990-2018)")
```

NOTE: I included this graph because I wasn't sure if we were allowed to use our visualization from the discussion post this week as a part of this assignment. I really wanted to show a bar or area chart that tracked the cumulative GHG production and compare it to worldwide production.